\documentclass{article}

% ---------------------------------------------------------------------------
% NeurIPS 2024 style. Download neurips_2024.sty from https://neurips.cc/Conferences/2024/PaperInformation/StyleFiles
% and place it next to this file. Use [preprint] while drafting; switch to the
% camera-ready/submission option at the venue we finally target.
% ---------------------------------------------------------------------------
\usepackage[preprint]{neurips_2024}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{xcolor}

% Draft note macro — strip before submission.
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\prelim}[1]{\textcolor{blue}{#1}}

\title{Disentangling Statistical Surprisal from Hierarchical Structure\\
in LLM--Brain Alignment}

\author{%
  Sudarssan N.\thanks{Correspondence: sudarssan73@gmail.com} \\
  \texttt{} \\
  % Add affiliations / co-authors here.
}

\begin{document}

\maketitle

\begin{abstract}
Large language model (LLM) hidden states linearly predict human language-network fMRI
responses at close to the noise ceiling, but it remains contested whether this alignment
reflects shared \emph{linguistic structure} or merely shared sensitivity to next-word
predictability (surprisal). We quantify, on the Pereira et al.\ (2018) benchmark, how much
voxel-level alignment is \emph{uniquely} explained by LLM hidden representations \emph{beyond}
surprisal, using cross-validated ridge encoding combined with variance partitioning
(commonality analysis). Across six models spanning $124$M--$500$M parameters and four model
families, hidden representations uniquely explain roughly two orders of magnitude more brain
variance than surprisal, and this advantage \emph{grows} with network depth. Contrasting
models, we find that the unique representational contribution does \emph{not} increase
monotonically with parameter count---larger models are more \emph{predictive} but not more
\emph{uniquely structured}---and that instruction tuning slightly \emph{reduces} both overall
alignment and unique-beyond-surprisal contribution. Our results, obtained with inference-only
compute, sharpen the structure-versus-statistics debate into a quantitative, reproducible
encoding-model question. \prelim{(Numbers in this draft are preliminary, pre--confidence-interval re-run.)}
\end{abstract}

\section{Introduction}
\label{sec:intro}

A central finding of contemporary neuro-AI is that the internal representations of transformer
language models predict neural responses in the human language network with striking accuracy:
\citet{schrimpf2021} report that GPT-2-xl predicts the Pereira2018 and Fedorenko2016 datasets
``at close to 100\% of the noise ceiling,'' with intermediate layers most predictive. As this
``predictivity'' has saturated, attention has shifted from \emph{whether} models align to
\emph{why} they do.

A live debate concerns whether alignment reflects shared linguistic \emph{structure} or shared
sensitivity to next-word \emph{predictability} (surprisal). \citet{kauf2023} argue that
lexical-semantic content, rather than syntactic form, drives ANN--brain similarity; other work
shows that highly capable models can become \emph{less} human-like on prediction-based
behavioural measures \citep{superhuman2025}. Surprisal and structural complexity are, however,
notoriously collinear, so a correlational answer is insufficient.

We take a deliberately quantitative stance. Rather than asking whether structure \emph{or}
surprisal explains alignment, we ask: \textbf{how much of LLM--brain alignment on Pereira2018 is
\emph{uniquely} explained by hidden representations \emph{beyond} surprisal, and how does that
residual depend on model size, layer depth, and instruction tuning?} We answer it with variance
partitioning over a voxel-wise encoding model, which cleanly separates the variance unique to
hidden states, unique to surprisal, and shared between them---robust to their collinearity.

\paragraph{Contributions.}
\begin{itemize}
  \item A reproducible, inference-only pipeline that reproduces the Schrimpf-style
  noise-ceiling-normalized predictivity curve and augments it with explicit surprisal control
  via variance partitioning (Section~\ref{sec:methods}).
  \item Evidence that hidden representations uniquely explain $\sim$50--70$\times$ more brain
  variance than surprisal, with the gap growing across network depth (Section~\ref{sec:core}).
  \item A model-family comparison showing that unique-beyond-surprisal contribution does
  \emph{not} grow monotonically with size, and that instruction tuning slightly reduces it
  (Sections~\ref{sec:size}--\ref{sec:instruct}).
\end{itemize}

\section{Related Work}
\label{sec:related}

\paragraph{LLM--brain predictivity.} \citet{schrimpf2021} established strong linear predictivity
of language-network responses from transformer activations; follow-up work related predictivity
to next-word prediction performance and to scale. \citet{merlin2022} controlled for changes in
next-word prediction and argued some alignment depends on information beyond it.

\paragraph{What drives alignment.} \citet{kauf2023} perturb word order and content/function
words on Pereira2018, concluding lexical-semantic content dominates; their analysis does not
statistically control for surprisal in the hidden-state-to-brain mapping nor construct
surprisal-matched sets. Joint information-theoretic and syntactic predictors of neural data have
been studied with MEG and naturalistic listening, showing structure-sensitive effects can
survive surprisal controls.

\paragraph{Our position.} We isolate the \emph{unique} voxel-level variance explained by hidden
states beyond surprisal within one encoding framework, and connect the residual to size, depth,
and instruction tuning---a combination not, to our knowledge, previously reported together.

\section{Methods}
\label{sec:methods}

\paragraph{Data.} We use the Pereira et al.\ (2018) language-network fMRI dataset
\citep{pereira2018}, accessed through the \texttt{brain-score/language} harness
\citep{schrimpf2021}. We analyse the $384$-sentence experiment ($9$ subjects). The assembly is
already restricted to language-network functional ROIs. Voxels are subject/experiment specific;
we drop voxels with missing coverage, yielding $384$ sentences $\times\ 12{,}155$ voxels.

\paragraph{Models.} We probe six autoregressive models across four families and a
$3\times$ size range: GPT-2 ($124$M) \citep{radford2019}, OPT-125M \citep{zhang2022opt},
Pythia-160M and Pythia-410M \citep{biderman2023pythia}, and the Qwen2.5-0.5B base/instruct pair
\citep{qwen2025}. All are run inference-only.

\paragraph{Activation extraction and pooling.} For each sentence we extract hidden states from
every layer and pool tokens to a single per-sentence vector using the \emph{last-token}
representation (a causal LM's final token has attended to the full sentence). We verified that
mean pooling collapses layer differences, consistent with a bag-of-words confound.

\paragraph{Surprisal features.} We compute per-token surprisal $-\log_2 P(t \mid t_{<i})$ from
the same model and summarize each sentence by its mean, sum, and final-token surprisal,
yielding a $3$-dimensional predictor.

\paragraph{Encoding model.} For features $X$ (activations or surprisal) and responses $Y$ we fit
$5$-fold cross-validated ridge regression with the regularization strength selected by efficient
leave-one-out generalized cross-validation, reporting per-voxel held-out Pearson $r$.

\paragraph{Noise ceiling.} Because Pereira voxels are subject-specific with no within-subject
repeats, we estimate a \emph{cross-subject} ceiling: each held-out subject's voxels are
predicted from the other subjects' responses to the same sentences (PCA-reduced cross-validated
ridge). Normalized predictivity is $r / \text{ceiling}$, averaged over voxels whose ceiling
exceeds $0.1$. \todo{state mean ceiling = 0.219 once finalized.}

\paragraph{Variance partitioning.} Given hidden features $H$ and surprisal features $S$, we fit
encoding models for $H$, $S$, and $H \cup S$ and decompose explained variance:
\begin{align}
\text{unique}(H) &= R^2(H \cup S) - R^2(S), &
\text{unique}(S) &= R^2(H \cup S) - R^2(H), \\
\text{shared} &= R^2(H) + R^2(S) - R^2(H \cup S). & &
\end{align}
We report $95\%$ confidence intervals by bootstrap over voxels.

\section{Results}
\label{sec:results}

\subsection{Reproducing noise-ceiling predictivity}
\label{sec:repro}
With last-token pooling and the cross-subject ceiling, GPT-2 reproduces the Schrimpf-style
result: normalized predictivity rises monotonically with depth to $\sim$$1.0$ at the final
layer (\prelim{$0.14 \rightarrow 1.02$}). For this $124$M model the peak is at the last layer;
the largest model (Pythia-410M) peaks at an intermediate layer, consistent with the reported
``peak moves earlier with scale.''

\subsection{Hidden representations dominate surprisal}
\label{sec:core}
Variance partitioning shows that hidden representations uniquely explain far more brain variance
than surprisal at every layer, and the advantage grows with depth. For GPT-2, peak
$\text{unique}(H) \approx \prelim{0.039}$ versus $\text{unique}(S) \approx \prelim{0.0006}$, a
ratio of $\sim$$65\times$. The effect holds across all six models (Table~\ref{tab:sweep}).

\subsection{Size: bigger is more predictive, not more uniquely structured}
\label{sec:size}
Table~\ref{tab:sweep} dissociates two notions of ``brain-like.'' Normalized predictivity tends
to rise with model quality/size (highest for Qwen2.5-0.5B). However, the \emph{unique}
contribution of hidden states beyond surprisal does \emph{not} increase monotonically with
parameter count---e.g.\ Pythia-410M does not exceed Pythia-160M. Scaling appears to buy
predictivity, not unique-beyond-surprisal structure.

\subsection{Instruction tuning slightly reduces alignment}
\label{sec:instruct}
For the Qwen2.5-0.5B base/instruct pair (matched architecture and size), instruction tuning
slightly \emph{reduces} both normalized predictivity (\prelim{$1.145 \rightarrow 1.127$}) and
peak unique-hidden (\prelim{$0.0489 \rightarrow 0.0467$}), both peaking at the same layer.
\todo{report paired bootstrap CI from Stage 7 (significant / n.s.).}

\begin{table}[t]
\caption{Model sweep on Pereira2018 ($384$-sentence experiment, last-token pooling,
cross-subject noise ceiling). \prelim{Preliminary; to be refreshed after the
CI/stabilized-ridge re-run.} ``norm $r$'' is best-layer normalized predictivity;
``unique$(H)$'' is peak across layers; ``unique$(S)$'' is at that layer.}
\label{tab:sweep}
\centering
\begin{tabular}{lrrrrr}
\toprule
Model & Params & best norm $r$ & peak unique$(H)$ & unique$(S)$ & $H{:}S$ \\
\midrule
GPT-2            & 124M & 1.02 & 0.039 & 0.0006 & 65$\times$ \\
OPT-125M         & 125M & 0.97 & 0.032 & 0.0005 & 61$\times$ \\
Pythia-160M      & 160M & 0.83 & 0.029 & 0.0006 & 48$\times$ \\
Pythia-410M      & 410M & 0.94 & 0.026 & \todo{} & \todo{} \\
Qwen2.5-0.5B     & 500M & 1.14 & 0.049 & 0.0008 & 58$\times$ \\
Qwen2.5-0.5B-Inst& 500M & 1.13 & 0.047 & 0.0009 & 54$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\section{Discussion and Limitations}
\label{sec:discussion}
Our central result---that hidden representations uniquely explain far more brain variance than
surprisal, growing with depth---supports a structural, rather than purely predictive, account of
LLM--brain alignment, while quantifying the (small but non-zero) unique surprisal contribution.

\paragraph{Limitations.} (i) We report a single experiment ($384$ sentences); the $243$-sentence
set should be added for robustness. (ii) Our cross-subject ceiling is principled but not
identical to the brain-score extrapolation ceiling; some normalized values slightly exceed $1$,
indicating a conservative ceiling estimate. (iii) Confidence intervals are bootstrap over
voxels; a stimulus-level bootstrap is stronger and planned. (iv) Encoding pools all
language-network voxels jointly rather than per subject. (v) Surprisal is a low-dimensional
summary; richer information-theoretic and symbolic-syntactic predictors would sharpen the
``structure'' side.

\section{Conclusion}
\label{sec:conclusion}
Framing the structure-versus-statistics debate as a variance-partitioning question yields a
clear, reproducible answer on Pereira2018: alignment is overwhelmingly carried by hidden
representations beyond surprisal, this is not a simple consequence of scale, and instruction
tuning does not help. The full pipeline runs inference-only and is released for reproduction.

\section*{Reproducibility}
Code, configuration, and experiment tracker are available at the project repository. All results
use public models and the public Pereira2018 release via \texttt{brain-score/language}.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
